---
knit: bookdown::preview_chapter
---

# Expression overview

## Introduction

Here we will continue to work with the filtered __blischak__ dataset produced in the previous chapter. We will look at what happened to the expression matrix after the quality control step.

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(out.width='90%', fig.align = 'center')
```

```{r, message=FALSE, warning=FALSE}
library(scater, quietly = TRUE)
options(stringsAsFactors = FALSE)
umi <- readRDS("blischak/umi.rds")
umi.qc <- umi[fData(umi)$use, pData(umi)$use]
endog_genes <- !fData(umi.qc)$is_feature_control
```

## PCA plot

The easiest thing to overview the data is to transform it using the principal component analysis and then visualize the first two principal components. Again the [scater](https://github.com/davismcc/scater) package provides several very useful functions to make this analysis straightforward.

### Before QC

```{r expr-overview-pca-before-qc, fig.cap = "PCA plot of the blischak data"}
scater::plotPCA(umi[endog_genes, ],
                colour_by = "batch",
                size_by = "total_features",
                exprs_values = "counts")
```

### After QC

```{r expr-overview-pca-after-qc, fig.cap = "PCA plot of the blischak data"}
scater::plotPCA(umi.qc[endog_genes, ],
                colour_by = "batch",
                size_by = "total_features",
                exprs_values = "counts")
```

## tSNE map

Another way of representing the data is a tSNE map. _We need to explain how tSNE works[dimensionality reduction + nearest neightbour network] & show the effect of changing the parameters (Also these plots look to me like the perplexity parameter is too low.)_

### Before QC

```{r expr-overview-tsne-before-qc, fig.cap = "tSNE map of the blischak data"}
scater::plotTSNE(umi[endog_genes, ],
                 colour_by = "batch",
                 size_by = "total_features",
                 exprs_values = "counts",
                 rand_seed = 123456)
```

### After QC

```{r expr-overview-tsne-after-qc, fig.cap = "tSNE map of the blischak data"}
scater::plotTSNE(umi.qc[endog_genes, ],
                 colour_by = "batch",
                 size_by = "total_features",
                 exprs_values = "counts",
                 rand_seed = 123456)
```

## Exercise

Perform the same analysis with read counts of the Blischak data. Use `blischak/reads.rds` file to load the reads SCESet object. Once you have finished please compare your results to ours (next chapter).
