---
knit: bookdown::preview_chapter
---

# Clustering analysis

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align = "center")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(scRNA.seq.funcs)
library(pcaMethods)
library(pcaReduce)
library(Rtsne)
library(SC3)
library(pheatmap)
library(ggplot2)
set.seed(1234567)
```

To illustrate clustering of scRNA-seq data, we consider a dataset of
hematopoietic stem cells (HSCs) collected from a patient with
[myeloproliferative neoplasm (MPN)](https://en.wikipedia.org/wiki/Myeloproliferative_neoplasm). It is well known that this disease is highly heterogeneous with multiple
sub-clones co-existing within the same patient. 

## Patient dataset

Traditionally, clonal heterogeneity has been assessed by genotyping
sub-clones. Genotyping through Sanger sequencing of two key loci has
been carried out for this patient and it has revealed the presence of
3 different sub-clones (WT, WT/Tet2 and Jak2/Tet2). Our goal is to
identify clusters corresponding to the three genotypes from the
scRNA-seq data.

```{r}
patient.data <- read.table("clustering/patient.txt", sep = "\t")
```

For convinience we have performed all quality control and normalization steps (SF + RUV) in advance. The
dataset contains `r dim(patient.data)[2]` cells and `r dim(patient.data)[1]` genes.

Here we present several methods available for clustering of scRNA-seq data.

## SC3

We first cluster this dataset using __SC3__:

```{r, eval=FALSE}
SC3::sc3(patient.data, ks = 2:5)
```

This command will open SC3 in a web browser. Once it is opened please perform the following exercises:

* __Exercise 1__: Explore different clustering solutions for $k$ from 2
to 5. Also try to change the consensus averaging by checking and
unchecking distance and transformation check boxes in the left panel
of __SC3__.

* __Exercise 2__: Based on the genotyping, we strongly believe that $k
=3$ provides the best clustering. What evidence do you find for this
in the clustering? How can you use the [silhouette plots](https://en.wikipedia.org/wiki/Silhouette_%28clustering%29) to
motivate choosing the value of $k$ that you think looks best?

* __Exercise 3__: The solution may change depending on the combination of
check-boxes that was used. How similar are the different solutions,
and which are the most stable clusterings? You can find out how
different cells migrate between clusters using the "Cell Labels" tab
panel.

* __Exercise 4__: Calculate differentially expressed genes and marker genes for the obtained clusterings. Please use $k=3$ and the most stable solution, corresponding to the
case when all check boxes are checked.

* __Exercise 5__: Change the marker genes threshold (the default is 0.85) in the right side panel of __SC3__. Does __SC3__ find more marker genes?

## pcaReduce

To run [pcaReduce](https://github.com/JustinaZ/pcaReduce) on the same dataset, use the following commands

```{r clust-pca-reduce3, fig.cap = "Clustering solutions of pcaReduce method after running it for 10 times and selecting $k=3$."}
# use the same gene filter as in SC3
input <- scRNA.seq.funcs::gene_filter(patient.data, 0.06)
# log transform before the analysis
input.log <- log2(input + 1)
# run pcaReduce 10 times creating hierarchies from 1 to 30 clusters
pca.red <- pcaReduce::PCAreduce(t(input.log), nbt = 10, q = 30, method = 'S')
res <- unique(scRNA.seq.funcs::merge_pcaReduce_results(pca.red, 3))
pheatmap::pheatmap(res)
```

__Exercise 6__: Run pcaReduce for $k=2$, $k=4$ and $k=5$. Is it easy to choose the optimal $k$?

__Hint__: When running pcaReduce for different $k$s you do not need to rerun pcaReduce::PCAreduce function, just use already calculated `pca.red` object.

__Our solutions__:
```{r clust-pca-reduce2, fig.cap = "Clustering solutions of pcaReduce method after running it for 10 times and selecting $k=2$.", echo=FALSE}
res <- unique(scRNA.seq.funcs::merge_pcaReduce_results(pca.red, 2))
pheatmap::pheatmap(res)
```

```{r clust-pca-reduce4, fig.cap = "Clustering solutions of pcaReduce method after running it for 10 times and selecting $k=4$.", echo=FALSE}
res <- unique(scRNA.seq.funcs::merge_pcaReduce_results(pca.red, 4))
pheatmap::pheatmap(res)
```

```{r clust-pca-reduce5, fig.cap = "Clustering solutions of pcaReduce method after running it for 10 times and selecting $k=5$.", echo=FALSE}
res <- unique(scRNA.seq.funcs::merge_pcaReduce_results(pca.red, 5))
pheatmap::pheatmap(res)
```

__Exercise 7__: Compare the results between SC3 and pcaReduce. What is
the main difference between the solutions provided by the two
different methods?

## tSNE + kmeans

[tSNE](https://lvdmaaten.github.io/tsne/) plots that we saw before (\@ref(visual-tsne)) when used the __scater__ package are made by using the [Rtsne](https://cran.r-project.org/web/packages/Rtsne/index.html) and [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html) packages. We can create a similar plots explicitly:
```{r clust-tsne, fig.cap = "tSNE map of the patient data"}
tsne_out <- Rtsne::Rtsne(t(input.log), perplexity = 10)
df_to_plot <- as.data.frame(tsne_out$Y)
comps <- colnames(df_to_plot)[1:2]
ggplot2::ggplot(df_to_plot, aes_string(x = comps[1], y = comps[2])) +
    geom_point() +
    xlab("Dimension 1") +
    ylab("Dimension 2") +
    theme_bw()
```

Note that all points on the plot above are black. This is different from what we saw before, when the cells were coloured based on the annotation. Here we do not have any annotation and all cells come from the same batch, therefore all dots are black.

Now we are going to apply _k_-means clustering algorithm to the cloud of points on the tSNE map. Do you see 2/3/4/5 groups in the cloud?

We will start with $k=2$:
```{r clust-tsne-kmeans2, fig.cap = "tSNE map of the patient data with 2 colored clusters, identified by the _k_-means clustering algorithm"}
clusts <- stats::kmeans(tsne_out$Y,
                        centers = 2,
                        iter.max = 1e+09,
                        nstart = 1000)$clust
df_to_plot$clusts <- factor(clusts, levels = unique(clusts))
comps <- colnames(df_to_plot)[1:3]
ggplot2::ggplot(df_to_plot,
                aes_string(x = comps[1], y = comps[2], color = comps[3])) +
    geom_point() +
    xlab("Dimension 1") +
    ylab("Dimension 2") +
    theme_bw()
```

__Exercise 7__: Make the same plots for $k=3$, $k=4$ and $k=5$.

__Exercise 8__: Compare the results between SC3 and tSNE+kmeans. Can the
results be improved by changing the perplexity parameter?

As you may have noticed, both pcaReduce and tSNE+kmeans are stochastic
and give different results every time they are run. To get a better
overview of the solutions, we need to run the methods multiple times. Here
we run __tSNE+kmeans__ clustering 10 times with $k = 3$ (with perplexity =
10):

```{r}
tsne.res <- scRNA.seq.funcs::tsne_mult(input.log, 3, 10)
res <- unique(do.call(rbind, tsne.res))
```

__Exercise 9__: Visualize the different clustering solutions using a
heatmap. Then run tSNE+kmeans algorithm with $k = 2$ or $k = 4$ and
see how the clustering looks like in these cases.

```{r clust-tsne-kmeans3, fig.cap = "Clustering solutions of tSNE+kmeans method after running it for 10 times and using $k=3$", echo=FALSE}
pheatmap::pheatmap(res, show_colnames = FALSE)
```

```{r clust-tsne-kmeans2-mult, fig.cap = "Clustering solutions of tSNE+kmeans method after running it for 10 times and using $k=2$", echo=FALSE}
tsne.res <- scRNA.seq.funcs::tsne_mult(input.log, 2, 10)
res <- unique(do.call(rbind, tsne.res))
pheatmap::pheatmap(res, show_colnames = FALSE)
```

```{r clust-tsne-kmeans4, fig.cap = "Clustering solutions of tSNE+kmeans method after running it for 10 times and using $k=4$", echo=FALSE}
tsne.res <- scRNA.seq.funcs::tsne_mult(input.log, 4, 10)
res <- unique(do.call(rbind, tsne.res))
pheatmap::pheatmap(res, show_colnames = FALSE)
```

## SNN-Cliq

To run the [SNN-Cliq](http://bioinfo.uncc.edu/SNNCliq/) method:

```{r}
distan <- "euclidean"
par.k <- 3
par.r <- 0.7
par.m <- 0.5
# construct a graph
scRNA.seq.funcs::SNN(
    data = t(input.log),
    outfile = "snn-cliq.txt",
    k = par.k,
    distance = distan
)
# find clusters in the graph
snn.res <- 
    system(
        paste0(
            "python ~/snn-cliq/Cliq.py ", 
            "-i snn-cliq.txt ",
            "-o res-snn-cliq.txt ",
            "-r ", par.r,
            " -m ", par.m
        ),
        intern = TRUE
    )
cat(paste(snn.res, collapse = "\n"))
snn.res <- read.table("res-snn-cliq.txt")
# remove files that were created during the analysis
system("rm snn-cliq.txt res-snn-cliq.txt")
```

__Exercise 10__: How can you characterize the solution identified by SNN-Cliq? Run SNN-Cliq algorithm with different values of _k_, _r_, _m_ and _distance_, and see how the clustering looks like in these cases.

## SINCERA

As mentioend in the previous chapter [SINCERA](https://research.cchmc.org/pbge/sincera.html) is based on hierarchical clustering. One important thing to keep in mind is that it performs a gene-level z-score transformation before doing clustering:

```{r, echo=TRUE, fig.height=7, fig.width=7}
# perform gene-by-gene per-sample z-score transformation
dat <- apply(input, 1, function(y) scRNA.seq.funcs::z.transform.helper(y))
# hierarchical clustering
dd <- as.dist((1 - cor(t(dat), method = "pearson"))/2)
hc <- hclust(dd, method = "average")
```

If the number of cluster is not known [SINCERA](https://research.cchmc.org/pbge/sincera.html) can identify __k__ as the minimum height of the hierarchical tree that generates no more than a specified number of singleton clusters (clusters containing only 1 cell)
```{r, echo=TRUE}
num.singleton <- 0
kk <- 1
for (i in 2:dim(dat)[2]) {
    clusters <- cutree(hc, k = i)
    clustersizes <- as.data.frame(table(clusters))
    singleton.clusters <- which(clustersizes$Freq < 2)
    if (length(singleton.clusters) <= num.singleton) {
        kk <- i
    } else {
        break;
    }
}
cat(kk)
```

__Exercise 11__: Visualize the SINCERA results as a heatmap. How do the
results compare to the other methods? What happens if you choose $k = 3$?

__Our answer:__
```{r clust-sincera, fig.cap = "Clustering solutions of SINCERA method using $k=3$", echo=FALSE}
pheatmap::pheatmap(t(dat),
                   cluster_cols = hc,
                   cutree_cols = 3,
                   kmeans_k = 100,
                   show_rownames = FALSE)
```


__Exercise 12__: Is using the singleton cluster criteria for finding __k__ a good idea?
